---
title: "Teaching AI the laws of physics via the principle of least action"
format: 
  html:
    toc: true
    theme: cosmo
---

## Introduction: two ways to look at a missile launch

Imagine a missile launched into the air. How does Nature decide exactly what path it will take before it hits the ground?

1.  **The Newtonian approach (local):** Nature looks at the missile frame-by-frame. At every single microsecond, it calculates the forces acting on it ($F = ma$) and pushes it to its next position.
2.  **The Lagrangian approach (global):** Nature looks at the *entire flight history all at once*. It looks at the starting point, the ending point, and chooses the path that minimizes a physical "cost function" called **Total Action**.

In this notebook, we are going to build a **neural network** that knows absolutely zero physics. It doesn't know what a parabola is, and it doesn't know Newton's equations. We will force it to search through an infinite space of possible curves, using optimization to discover the exact path of a missile by minimizing its Action cost.

## Defining the "cost" of a trajectory

In physics, the cost function at any given moment is called the **Lagrangian (**$L$). It is simply the difference between Kinetic Energy ($T$) and Potential Energy ($V$):

$$L = T - V$$

For a missile of mass $m$ under standard gravity $g$:

- **Kinetic Energy (Motion Cost):** $T = \frac{1}{2}m(v_x^2 + v_y^2)$

- **Potential Energy (Height Cost):** $V = mgy$

The **Total Action** ($S$) is the sum of this cost across the entire flight time. Nature always picks the path where $S$ is minimized:

$$S = \int_{0}^{t_{max}} (T - V) \, dt$$

If our neural network guesses a path that is too wavy, the velocity spikes, making $T$ huge, and $S$ explodes. If it guesses a path that is a flat line, it misses out on the potential energy deductions from $V$, and $S$ increases. The perfect balance is a parabola.

## Fit neural network to find "the path of least action"

We will write a tiny neural network from scratch using base R.

- **Input (**$t$): A specific time.
- **Output (**$y$): The predicted altitude of the missile at that time.
- **Loss Function:** The Total Action ($S$) + a massive penalty if the network fails to start at $y=0$ or land at $y=0$.

```{r setup}
# Initialize random seed for reproducibility
set.seed(42)

# 1. Physical Constants
m <- 1.0     # Missile mass (kg)
g <- 9.81    # Gravity (m/s^2)
t_max <- 2   # Total flight time limit (seconds)
vx <- 10     # Fixed horizontal speed (travels 20m in 2 seconds)
```

### Setup network and loss function

The code below constructs the forward pass of the network and sets up the calculus loop that approximates the Total Action integral.

```{r nn-engine}
# 2. A Simple Multi-Layer Perceptron Function
# 1 Input (time) -> 5 Hidden Neurons (with smooth tanh curve) -> 1 Output (position y)
neural_network_path <- function(t, weights) {
  w1 <- weights[1:5]
  b1 <- weights[6:10]
  w2 <- weights[11:15]
  b2 <- weights[16]
  
  # Forward pass math: activation = tanh(t * w1 + b1)
  hidden <- tanh(t %*% t(w1) + matrix(b1, nrow=length(t), ncol=5, byrow=TRUE))
  y <- hidden %*% w2 + b2
  return(as.vector(y))
}

# 3. The Custom Physics Loss Function
nn_loss_function <- function(weights) {
  dt <- 0.01
  t_grid <- seq(0, t_max, by = dt)
  
  # Get NN predicted positions across the timeline
  y <- neural_network_path(t_grid, weights)
  
  # Calculate velocity (dy/dt) using numerical finite differences
  vy <- diff(y) / dt
  y_trimmed <- y[-length(y)]
  
  # Energy equations
  T_eng <- 0.5 * m * (vx^2 + vy^2)
  V_eng <- m * g * y_trimmed
  L <- T_eng - V_eng
  
  # Total Action (S) via Riemann sum approximation
  action_S <- sum(L * dt)
  
  # Boundary Constraints: The network MUST start and end at ground level (y = 0)
  y_start <- neural_network_path(0, weights)
  y_end   <- neural_network_path(t_max, weights)
  boundary_penalty <- 10000 * (y_start^2 + y_end^2)
  
  return(action_S + boundary_penalty)
}
```

### Train network

Now, we feed a set of random weights into the network. It will look like a chaotic scribble. Then, we let R's optimizer adjust those weights to minimize the Total Action cost.

```{r training}
# 4. Initialize Network Weights Randomly
initial_weights <- runif(16, min = -0.5, max = 0.5)

# 5. Optimize the Weights using Gradient Descent (BFGS algorithm)
nn_optimization <- optim(
  par = initial_weights, 
  fn = nn_loss_function, 
  method = "BFGS", 
  control = list(maxit = 2000, reltol = 1e-8)
)

trained_weights <- nn_optimization$par
```

## Compare with trajectory calculated using Newtonian mechanics

To see if this variational deep-learning approach actually works, let's overlay the neural network's self-discovered path against the exact calculus path calculated using Newtonian mechanics: $y(t) = v_{0y}t - \frac{1}{2}gt^2$.

```{r plotting, fig.width=8, fig.height=5}
# 6. Generate Paths for the Final Graph
t_plot <- seq(0, t_max, by = 0.01)

# Paths generated by our trained AI
y_nn <- neural_network_path(t_plot, trained_weights)

# Paths derived from textbook physics
v0y <- 0.5 * g * t_max
y_true <- (v0y * t_plot) - (0.5 * g * t_plot^2)

# 7. Render Plot
plot(t_plot, y_true, type = "l", col = "orange", lwd = 6,
     xlab = "Time (seconds)", ylab = "Altitude (meters)",
     main = "AI Self-Discovery of Projectile Motion via Least Action")
lines(t_plot, y_nn, col = "blue", lwd = 2, lty = 2)
grid()
legend("topright", legend = c("Exact Newtonian Path", "Neural Network Learned Path"),
       col = c("orange", "blue"), lty = c(1, 2), lwd = c(6, 2))
```

------------------------------------------------------------------------

## Takeaways for data scientists

What just happened here?

- **No physics equations were harmed:** We never gave the network the equation of a parabola ($t^2$). We never gave it an acceleration formula.
- **Emergent behavior:** The network discovered a perfect parabola purely because *any other shape would cost more energy*.
- **Why this matters:** This is the underlying concept behind **Physics-Informed Neural Networks (PINNs)**. When physical systems (like fluid dynamics, climate modeling, or quantum mechanics) become too complex to solve using Newton's step-by-step vector equations, we can turn the problem into a global geometry search using neural networks.
