---
title: "KL Divergence Minimization as a General Principle of Statistical Learning"
subtitle: "From Maximum Likelihood to Bayesian Deep Learning"
format:
  html:
    toc: true
    toc-depth: 3
    code-fold: show
execute:
  warning: false
  message: false
---

## Overview and learning objectives

Most statistical and machine learning methods are introduced through disparate lenses: Ordinary Least Squares minimizes squared residuals, Maximum Likelihood maximizes sample probability density, and Bayesian inference updates prior beliefs via Bayes' rule.

This notebook demonstrates that **Kullback-Leibler (KL) divergence minimization serves as a unifying foundational principle across all these paradigms**:

1.  **Maximum Likelihood Estimation (MLE)** minimizes the *forward KL divergence in data space* between the true distribution and the model family.
2.  **Mean Squared Error (MSE)** is the exact mathematical consequence of minimizing this data-space KL divergence under Gaussian errors.
3.  **Bayesian Learning (Variational Inference)** minimizes the *reverse KL divergence in parameter space* between an approximate posterior distribution and the true posterior distribution.

## Sources and further reading

- [VIB is Half Bayes](https://youtu.be/JGDAZ4joUX8) [VIB is Half Bayes (arXiv:2011.08711)](https://doi.org/10.48550/arXiv.2011.08711)
- [A Tale of Two Worlds: The Variational Approach to Machine Learning](https://youtu.be/OAHsR02dUc0)

```{r}
#| label: setup
library(tidyverse)
library(ggformula)
library(torch)

# Set seeds for reproducibility across R and libtorch
set.seed(666)
torch_manual_seed(666)
```

## The true data-generating process (DGP)

We define a known synthetic ground-truth environment $p^*(x, y) = p^*(y \mid x) p^*(x)$:

$$X \sim \text{Uniform}(-5, 5)$$ $$Y \mid X \sim \text{Normal}\left(\mu^*(X) = \frac{X^3}{100}, \; \sigma^{*2} = 0.09\right)$$

The true conditional standard deviation is $\sigma^* = \sqrt{0.09} = 0.3$. We sample a small dataset of $n = 30$ observations to illustrate how different learning principles behave in the low-data / high-capacity regime.

```{r}
#| label: simulate-data
# Sample size
n <- 30

# True DGP parameters
sigma_sq_true <- 0.09
sigma_true    <- sqrt(sigma_sq_true)

# Simulated training sample
sim_data <- tibble(
  x = runif(n, min = -5, max = 5)
) |>
  mutate(
    mu_true = (x^3) / 100,
    y = rnorm(n, mean = mu_true, sd = sigma_true)
  )

# Dense evaluation grid for the ground truth
true_grid <- tibble(x = seq(-5, 5, length.out = 300)) |>
  mutate(
    mu_true = (x^3) / 100,
    lower = mu_true - sigma_true,
    upper = mu_true + sigma_true
  )

# Visualize DGP
gf_point(y ~ x, data = sim_data, alpha = 0.6, size = 2, color = "gray20") |>
  gf_ribbon(lower + upper ~ x, data = true_grid, fill = "peachpuff", alpha = 0.6, inherit = FALSE) |>
  gf_line(mu_true ~ x, data = true_grid, color = "darkorange", linetype = "dashed", linewidth = 1, inherit = FALSE) |>
  gf_lims(y = c(-3, 3), x = c(-6, 6)) |>
  gf_labs(
    title = "True Data Distribution & Sampled Observations (n = 30)",
    subtitle = "Dashed orange line: true mean function; Peach envelope: true ±1 SD",
    x = "X",
    y = "Y"
  ) |>
  gf_theme(theme_minimal())
```

## Theory: learning via data-space KL divergence minimization

Let $p^*(x, y)$ be the true distribution and $q_\theta(x, y) = q_\theta(y \mid x) p^*(x)$ be our parameterized model candidate.

The **Kullback-Leibler (KL) divergence** from our candidate model $q_\theta$ to the true distribution $p^*$ measures information lost when approximating $p^*$ with $q_\theta$:

$$D_{\text{KL}}(p^* \parallel q_\theta) = \int \int p^*(x, y) \log \frac{p^*(x, y)}{q_\theta(x, y)} \, dy \, dx$$

Expanding the logarithm:

$$D_{\text{KL}}(p^* \parallel q_\theta) = \underbrace{\mathbb{E}_{p^*}\left[\log p^*(X, Y)\right]}_{\text{Constant with respect to } \theta} - \mathbb{E}_{p^*}\left[\log q_\theta(Y \mid X)\right]$$

Because the entropy of the real world $\mathbb{E}_{p^*}[\log p^*(X, Y)]$ is invariant to our model parameters $\theta$, minimizing KL divergence directly simplifies to:

$$\arg\min_\theta D_{\text{KL}}(p^* \parallel q_\theta) = \arg\max_\theta \mathbb{E}_{p^*}\left[\log q_\theta(Y \mid X)\right]$$

Under an observed finite sample $(x_i, y_i)_{i=1}^n \sim p^*$, the empirical sample mean approximates the population expectation:

$$\mathbb{E}_{p^*}\left[\log q_\theta(Y \mid X)\right] \approx \frac{1}{n}\sum_{i=1}^n \log q_\theta(y_i \mid x_i)$$

$$\therefore \arg\min_\theta D_{\text{KL}}(p^* \parallel q_\theta) \approx \arg\max_\theta \sum_{i=1}^n \log q_\theta(y_i \mid x_i)$$

> **Core Insight 1:** Maximum Likelihood Estimation (MLE) is not an arbitrary estimation heuristic; it is the empirical approximation of **forward KL divergence minimization in data space**.

### Deriving mean squared error (MSE) from the Gaussian log-likelihood

Assume a homoscedastic conditional Gaussian model with constant noise variance $\sigma^2$:

$$q_\theta(Y \mid X = x) = \frac{1}{\sqrt{2\pi\sigma^2}} \exp\left( -\frac{(y - f(x; \theta))^2}{2\sigma^2} \right)$$

Evaluating the total log-likelihood:

$$\sum_{i=1}^n \log q_\theta(y_i \mid x_i) = -\frac{n}{2}\log(2\pi\sigma^2) - \frac{1}{2\sigma^2}\sum_{i=1}^n \left(y_i - f(x_i; \theta)\right)^2$$

Dropping additive constants and positive scaling factors independent of $\theta$:

$$\arg\max_\theta \sum_{i=1}^n \log q_\theta(y_i \mid x_i) \iff \arg\min_\theta \frac{1}{n} \sum_{i=1}^n \left(y_i - f(x_i; \theta)\right)^2$$

> **Core Insight 2:** Ordinary Least Squares / Mean Squared Error regression is mathematically equivalent to empirical KL divergence minimization under a constant-variance Gaussian model.

## Polynomial hypotheses and the capacity dilemma

We first evaluate data-space KL minimization across linear regression models with three different hypothesis capacities:

1.  **Linear (Underparameterized / High Bias):** $f(x) = \beta_0 + \beta_1 x$
2.  **Cubic (Well-Specified):** $f(x) = \sum_{k=0}^3 \beta_k x^k$
3.  **Degree 9 Polynomial (Overparameterized / High Variance):** $f(x) = \sum_{k=0}^9 \beta_k x^k$

```{r}
#| label: model-fitting
mod_linear <- lm(y ~ x, data = sim_data)
mod_cubic  <- lm(y ~ poly(x, 3, raw = TRUE), data = sim_data)
mod_deg9   <- lm(y ~ poly(x, 9, raw = TRUE), data = sim_data)

preds_poly <- true_grid |>
  mutate(
    `Linear (Underfit)` = predict(mod_linear, newdata = true_grid),
    `Cubic (Well-Specified)` = predict(mod_cubic, newdata = true_grid),
    `Degree 9 (Overfit)` = predict(mod_deg9, newdata = true_grid)
  ) |>
  pivot_longer(
    cols = c(`Linear (Underfit)`, `Cubic (Well-Specified)`, `Degree 9 (Overfit)`),
    names_to = "Model",
    values_to = "y_hat"
  )

gf_point(y ~ x, data = sim_data, alpha = 0.5, size = 1.8, color = "gray25") |>
  gf_ribbon(lower + upper ~ x, data = true_grid, fill = "peachpuff", alpha = 0.5, inherit = FALSE) |>
  gf_line(mu_true ~ x, data = true_grid, color = "darkorange", linetype = "dashed", linewidth = 0.8, inherit = FALSE) |>
  gf_line(y_hat ~ x, data = preds_poly, color = ~Model, linewidth = 1, inherit = FALSE) |>
  gf_lims(y = c(-3, 3), x = c(-5, 5)) |>
  gf_labs(
    title = "Polynomial Hypothesis Classes via Least Squares (Data-Space KL)",
    subtitle = "Higher capacity models minimize sample empirical risk at the cost of divergence from p*",
    x = "X",
    y = "Y"
  ) |>
  gf_theme(theme_minimal())
```

## Analytical evaluation of generalization KL divergence

To quantify model quality, we compute the true population KL divergence on test inputs $X \sim \text{Uniform}(-5, 5)$.

For two Gaussians $p^*(Y \mid x) = \mathcal{N}(\mu^*(x), \sigma^{*2})$ and $q(Y \mid x) = \mathcal{N}(\hat{\mu}(x), \hat{\sigma}^2)$, the conditional KL divergence (in nats) is:

$$D_{\text{KL}}(p^*(\cdot \mid x) \parallel q(\cdot \mid x)) = \log\frac{\hat{\sigma}}{\sigma^*} + \frac{\sigma^{*2} + (\mu^*(x) - \hat{\mu}(x))^2}{2\hat{\sigma}^2} - \frac{1}{2}$$

Dividing by $\log(2)$ yields the divergence in **bits**.

```{r}
#| label: kl-poly-evaluation
# Dense test sample
test_data <- tibble(
  x = runif(20000, -5, 5),
  mu_true = (x^3) / 100
) |>
  mutate(
    pred_linear = predict(mod_linear, newdata = pick(everything())),
    pred_cubic  = predict(mod_cubic,  newdata = pick(everything())),
    pred_deg9   = predict(mod_deg9,   newdata = pick(everything()))
  )

# Function to compute conditional KL divergence in bits
compute_kl_bits <- function(mu_true, mu_pred, sigma_pred, sigma_true = 0.3) {
  kl_nats <- log(sigma_pred / sigma_true) + 
    (sigma_true^2 + (mu_true - mu_pred)^2) / (2 * (sigma_pred^2)) - 0.5
  kl_nats / log(2)
}

# Empirical residual standard deviations for polynomial fits
sigma_linear <- summary(mod_linear)$sigma
sigma_cubic  <- summary(mod_cubic)$sigma
sigma_deg9   <- summary(mod_deg9)$sigma

poly_results <- test_data |>
  summarise(
    `Linear (Underfit)`    = mean(compute_kl_bits(mu_true, pred_linear, sigma_linear)),
    `Cubic (Well-Specified)` = mean(compute_kl_bits(mu_true, pred_cubic,  sigma_cubic)),
    `Degree 9 (Overfit)`   = mean(compute_kl_bits(mu_true, pred_deg9,   sigma_deg9))
  ) |>
  pivot_longer(everything(), names_to = "Model", values_to = "Expected_KL_Bits")

knitr::kable(poly_results, digits = 4, caption = "Expected KL Divergence from True DGP in Bits (Lower is Better)")
```

## Neural networks via MLE: high capacity meets empirical risk

A deep neural network parameterized by $\theta$ acts as a universal function approximator. We fit a Multi-Layer Perceptron (MLP) using the same homoscedastic likelihood:

$$Y \mid X \sim \text{Normal}\left(\mu = f(X; \theta), \; \hat{\sigma}^2\right)$$

- **Architecture:** $1 \to 32 \to 32 \to 1$ (\~1,150 parameters)
- **Activations:** Continuously Differentiable Exponential Linear Units (`CELU`)
- **Loss:** Mean Squared Error (MSE), directly minimizing empirical data-space KL divergence.

```{r}
#| label: mle-neural-network
# Define Homoscedastic MLP Architecture
MLP <- nn_module(
  "MLP",
  initialize = function() {
    self$net <- nn_sequential(
      nn_linear(1, 32),
      nn_celu(),
      nn_linear(32, 32),
      nn_celu(),
      nn_linear(32, 1)
    )
  },
  forward = function(x) {
    self$net(x)
  }
)

mle_model <- MLP()
mle_optimizer <- optim_adam(mle_model$parameters, lr = 0.005)

# Tensors for torch
x_train <- torch_tensor(matrix(sim_data$x, ncol = 1), dtype = torch_float())
y_train <- torch_tensor(matrix(sim_data$y, ncol = 1), dtype = torch_float())

# Train network by minimizing MSE
mle_epochs <- 3000
for (epoch in 1:mle_epochs) {
  mle_optimizer$zero_grad()
  
  y_hat <- mle_model(x_train)
  loss  <- nnf_mse_loss(y_hat, y_train)
  
  loss$backward()
  mle_optimizer$step()
}

# Estimate constant residual standard deviation sigma_hat
mle_model$eval()
with_no_grad({
  residuals <- y_train - mle_model(x_train)
  sigma_hat_mle <- as.numeric(torch_sqrt(torch_mean(residuals^2)))
})
```

```{r}
#| label: mle-visualization
# Predict on an extended evaluation grid x in [-6, 6] to observe extrapolation
eval_grid <- tibble(x = seq(-6, 6, length.out = 400))
x_test_tensor <- torch_tensor(matrix(eval_grid$x, ncol = 1), dtype = torch_float())

with_no_grad({
  mu_mle <- as.numeric(mle_model(x_test_tensor))
})

mle_eval <- eval_grid |>
  mutate(
    mu_hat     = mu_mle,
    sigma_hat  = sigma_hat_mle,
    lower_band = mu_hat - sigma_hat,
    upper_band = mu_hat + sigma_hat
  )

# Calculate test KL divergence on support [-5, 5]
eval_mle_kl <- test_data |>
  mutate(
    pred_mle = approx(mle_eval$x, mle_eval$mu_hat, xout = x)$y,
    kl_bits  = compute_kl_bits(mu_true, pred_mle, sigma_hat_mle)
  )

mle_kl_bits <- round(mean(eval_mle_kl$kl_bits), 3)

# Visualization
gf_point(y ~ x, data = sim_data, alpha = 0.5, size = 1.8, color = "gray25") |>
  gf_ribbon(lower + upper ~ x, data = true_grid, fill = "peachpuff", alpha = 0.5, inherit = FALSE) |>
  gf_line(mu_true ~ x, data = true_grid, color = "darkorange", linetype = "dashed", linewidth = 0.8, inherit = FALSE) |>
  gf_ribbon(lower_band + upper_band ~ x, data = mle_eval, fill = "#1f77b4", alpha = 0.35, inherit = FALSE) |>
  gf_line(mu_hat ~ x, data = mle_eval, color = "#1f77b4", linewidth = 1.1, inherit = FALSE) |>
  gf_lims(y = c(-3, 3), x = c(-6, 6)) |>
  gf_labs(
    title = "Neural Network via MLE (Data-Space KL Minimization)",
    subtitle = paste0("Mean conditional KL divergence: ", mle_kl_bits, " bits (Estimated sigma: ", round(sigma_hat_mle, 3), ")"),
    x = "X",
    y = "Y"
  ) |>
  gf_theme(theme_minimal())
```

With 1,150 parameters and only 30 observations, empirical data-space KL minimization interpolates noise clusters and exhibits arbitrary behavior outside the training domain $x \notin [-5, 5]$.

## Bayesian learning: parameter-space KL divergence minimization

To tame overfitting, Bayesian inference avoids searching for a single optimal parameter point $\theta$. Instead, we place a prior distribution $p(\theta)$ over network weights and seek the posterior distribution $p(\theta \mid \mathcal{D})$.

Because the exact posterior is analytically intractable in deep networks, **Variational Inference** frames parameter learning as finding an approximate distribution $q_\phi(\theta)$ that minimizes the **KL divergence in parameter space**:

$$\phi^* = \arg\min_\phi D_{\text{KL}}\left(q_\phi(\theta) \;\parallel\; p(\theta \mid \mathcal{D})\right)$$

By Bayes' theorem, $p(\theta \mid \mathcal{D}) = \frac{p(\mathcal{D} \mid \theta) p(\theta)}{p(\mathcal{D})}$:

$$D_{\text{KL}}\left(q_\phi(\theta) \;\parallel\; p(\theta \mid \mathcal{D})\right) = \underbrace{D_{\text{KL}}\left(q_\phi(\theta) \;\parallel\; p(\theta)\right)}_{\text{Parameter-Space Complexity Penalty}} - \underbrace{\mathbb{E}_{q_\phi(\theta)}\left[\sum_{i=1}^n \log q_\theta(y_i \mid x_i)\right]}_{\text{Expected Data Log-Likelihood}} + \underbrace{\log p(\mathcal{D})}_{\text{Constant wrt } \phi}$$

Discarding the marginal evidence $\log p(\mathcal{D})$, minimizing this parameter-space KL divergence is identical to minimizing the **Variational Free Energy** (maximizing the Evidence Lower Bound, ELBO):

$$\mathcal{L}_{\text{VI}}(\phi) = D_{\text{KL}}\left(q_\phi(\theta) \;\parallel\; p(\theta)\right) + \sum_{i=1}^n \mathbb{E}_{q_\phi(\theta)}\left[ \frac{(y_i - f(x_i; \theta))^2}{2\hat{\sigma}^2} + \log \hat{\sigma} \right]$$

### The reparameterization trick

We specify an isotropic prior $p(\theta) = \mathcal{N}(0, \sigma_0^2 I)$ and a mean-field Gaussian variational posterior $q_\phi(w) = \mathcal{N}(\mu_w, \sigma_w^2)$ where $\sigma_w = 10^{-4} + \text{softplus}(\rho_w)$.

Weights are sampled differentiably via: $$w = \mu_w + \sigma_w \odot \epsilon, \quad \epsilon \sim \mathcal{N}(0, I)$$

```{r}
#| label: bayesian-nn-model
# Variational Bayesian Linear Layer
bnn_linear <- nn_module(
  "bnn_linear",
  initialize = function(in_features, out_features, prior_sd = 1.0) {
    self$in_features  <- in_features
    self$out_features <- out_features
    self$prior_sd     <- prior_sd

    # Variational parameters: means and log-variances (rho)
    self$weight_mu  <- nn_parameter(torch_empty(out_features, in_features)$uniform_(-0.1, 0.1))
    self$weight_rho <- nn_parameter(torch_full(c(out_features, in_features), -4.0))

    self$bias_mu  <- nn_parameter(torch_zeros(out_features))
    self$bias_rho <- nn_parameter(torch_full(c(out_features), -4.0))
  },
  forward = function(x) {
    weight_sigma <- 1e-4 + nnf_softplus(self$weight_rho)
    bias_sigma   <- 1e-4 + nnf_softplus(self$bias_rho)

    # Reparameterization trick
    eps_w <- torch_randn_like(self$weight_mu)
    eps_b <- torch_randn_like(self$bias_mu)

    weight <- self$weight_mu + weight_sigma * eps_w
    bias   <- self$bias_mu   + bias_sigma   * eps_b

    nnf_linear(x, weight, bias)
  },
  kl_loss = function() {
    weight_sigma <- 1e-4 + nnf_softplus(self$weight_rho)
    bias_sigma   <- 1e-4 + nnf_softplus(self$bias_rho)

    # Analytical KL against N(0, prior_sd^2)
    kl_w <- torch_sum(torch_log(self$prior_sd / weight_sigma) + 
                      (weight_sigma^2 + self$weight_mu^2) / (2 * self$prior_sd^2) - 0.5)
    kl_b <- torch_sum(torch_log(self$prior_sd / bias_sigma) + 
                      (bias_sigma^2 + self$bias_mu^2) / (2 * self$prior_sd^2) - 0.5)
    kl_w + kl_b
  }
)

# Complete Variational BNN Architecture
BayesianMLP <- nn_module(
  "BayesianMLP",
  initialize = function(prior_sd = 1.0) {
    self$l1 <- bnn_linear(1, 32, prior_sd = prior_sd)
    self$l2 <- bnn_linear(32, 32, prior_sd = prior_sd)
    self$l3 <- bnn_linear(32, 1, prior_sd = prior_sd)
  },
  forward = function(x) {
    x <- nnf_celu(self$l1(x))
    x <- nnf_celu(self$l2(x))
    self$l3(x)
  },
  kl_loss = function() {
    self$l1$kl_loss() + self$l2$kl_loss() + self$l3$kl_loss()
  }
)

bnn_model <- BayesianMLP(prior_sd = 1.0)
bnn_optimizer <- optim_adam(bnn_model$parameters, lr = 0.005)

# Variational Free Energy Optimization
bnn_epochs <- 3000
for (epoch in 1:bnn_epochs) {
  bnn_optimizer$zero_grad()
  
  # KL annealing schedule: beta scales smoothly from 0 to 1/n
  beta <- min(1 / n, (epoch / 1500) * (1 / n))
  
  # Stochastic forward pass: sample theta ~ q_phi(theta)
  y_hat <- bnn_model(x_train)
  
  # Negative log-likelihood (constant noise model with true/calibrated sigma)
  nll <- torch_sum(0.5 * torch_pow((y_train - y_hat) / sigma_true, 2))
  
  # Total Variational Free Energy: Data NLL + beta * Parameter KL
  loss <- nll + beta * bnn_model$kl_loss()
  
  loss$backward()
  bnn_optimizer$step()
}
```

### Predictive epistemic and aleatoric uncertainty

Under the Bayesian model, generating predictions entails drawing $S$ posterior parameter samples $\theta^{(s)} \sim q_\phi(\theta)$.

The total predictive variance decomposes into: $$\sigma^2_{\text{pred}}(x) = \underbrace{\text{Var}_{\theta}\left[f(x; \theta)\right]}_{\text{Epistemic (Model Uncertainty)}} + \underbrace{\sigma^{*2}}_{\text{Aleatoric (Noise)}}$$

```{r}
#| label: bnn-visualization
# Draw Monte Carlo predictions from the variational posterior
num_mc_samples <- 100
mc_preds <- matrix(0, nrow = nrow(eval_grid), ncol = num_mc_samples)

with_no_grad({
  for (s in 1:num_mc_samples) {
    mc_preds[, s] <- as.numeric(bnn_model(x_test_tensor))
  }
})

bnn_eval <- eval_grid |>
  mutate(
    mu_hat        = rowMeans(mc_preds),
    var_epistemic = apply(mc_preds, 1, var),
    var_aleatoric = sigma_sq_true,
    total_sd      = sqrt(var_epistemic + var_aleatoric),
    lower_band    = mu_hat - total_sd,
    upper_band    = mu_hat + total_sd
  )

# Calculate test KL divergence on support [-5, 5]
eval_bnn_kl <- test_data |>
  mutate(
    pred_bnn   = approx(bnn_eval$x, bnn_eval$mu_hat, xout = x)$y,
    sigma_bnn  = approx(bnn_eval$x, bnn_eval$total_sd, xout = x)$y,
    kl_bits    = compute_kl_bits(mu_true, pred_bnn, sigma_bnn)
  )

bnn_kl_bits <- round(mean(eval_bnn_kl$kl_bits), 3)

# Visualization
gf_point(y ~ x, data = sim_data, alpha = 0.5, size = 1.8, color = "gray25") |>
  gf_ribbon(lower + upper ~ x, data = true_grid, fill = "peachpuff", alpha = 0.5, inherit = FALSE) |>
  gf_line(mu_true ~ x, data = true_grid, color = "darkorange", linetype = "dashed", linewidth = 0.8, inherit = FALSE) |>
  gf_ribbon(lower_band + upper_band ~ x, data = bnn_eval, fill = "#2ca02c", alpha = 0.3, inherit = FALSE) |>
  gf_line(mu_hat ~ x, data = bnn_eval, color = "#2ca02c", linewidth = 1.1, inherit = FALSE) |>
  gf_lims(y = c(-3, 3), x = c(-6, 6)) |>
  gf_labs(
    title = "Bayesian Neural Network (Parameter-Space KL Minimization)",
    subtitle = paste0("Mean conditional KL divergence: ", bnn_kl_bits, " bits (Total predictive uncertainty: epistemic + aleatoric)"),
    x = "X",
    y = "Y"
  ) |>
  gf_theme(theme_minimal())
```

Notice how the Bayesian network's uncertainty band naturally flares outward for $|x| > 5$ where data is absent, accurately communicating epistemic uncertainty rather than making overconfident assertions.

## Synthesis: the hierarchy of KL divergence minimization

```{r}
#| label: final-comparison
all_results <- tibble(
  Paradigm = c(
    "Linear Regression (OLS)",
    "Cubic Regression (OLS)",
    "Degree 9 Polynomial (OLS)",
    "Neural Network (MLE / MSE)",
    "Bayesian Neural Network (Variational Inference)"
  ),
  `Optimization Objective` = c(
    "Min Data-Space KL (Linear Family)",
    "Min Data-Space KL (Cubic Family)",
    "Min Data-Space KL (Polynomial Deg 9)",
    "Min Data-Space KL (Neural Network)",
    "Min Parameter-Space KL (ELBO Free Energy)"
  ),
  `Expected KL Divergence (Bits)` = c(
    mean(compute_kl_bits(test_data$mu_true, test_data$pred_linear, sigma_linear)),
    mean(compute_kl_bits(test_data$mu_true, test_data$pred_cubic,  sigma_cubic)),
    mean(compute_kl_bits(test_data$mu_true, test_data$pred_deg9,   sigma_deg9)),
    mean(eval_mle_kl$kl_bits),
    mean(eval_bnn_kl$kl_bits)
  )
)

knitr::kable(all_results, digits = 4, caption = "Summary: KL Divergence Across Statistical Learning Paradigms")
```

### Summary of principles

| Paradigm | Domain of KL Minimization | Mathematical Formulation | Typical Pathology |
|:-----------------|:-----------------|:-----------------|:-----------------|
| **Maximum Likelihood (MLE / OLS)** | **Data Space:** $D_{\text{KL}}(p^* \parallel q_\theta)$ | $\arg\max_\theta \sum_{i=1}^n \log q_\theta(y_i \mid x_i)$ | Overfits and collapses uncertainty in high-capacity models |
| **Bayesian Inference (VI)** | **Parameter Space:** $D_{\text{KL}}(q_\phi \parallel p(\cdot \mid \mathcal{D}))$ | $\arg\min_\phi \left\{ D_{\text{KL}}(q_\phi \parallel p) - \mathbb{E}_{q}[\log q(y \mid x)] \right\}$ | Regularizes capacity and preserves honest epistemic uncertainty |
