---
title: "KL Divergence Minimization as a General Principle of Statistical Learning"
format: html
execute:
  warning: false
  message: false
---

```{r}
#| label: setup
library(tidyverse)
library(ggformula)

set.seed(42)
```

## Sources

https://youtu.be/JGDAZ4joUX8

https://doi.org/10.48550/arXiv.2011.08711


## Simulating the True Data Generating Process

From the target specification, the true distribution is defined as:

$$X \sim \text{Uniform}(-5, 5)$$
$$Y \mid X \sim \text{Normal}\left(\frac{X^3}{100}, 0.09\right)$$

Note that the variance is $\sigma^2 = 0.09$, which corresponds to a standard deviation of $\sigma = \sqrt{0.09} = 0.3$.

```{r}
#| label: simulate-data
# Number of samples
n <- 300

# True DGP parameters
sim_data <- tibble(
  x = runif(n, min = -5, max = 5)
) |>
  mutate(
    mu_true = (x^3) / 100,
    y = rnorm(n, mean = mu_true, sd = sqrt(0.09))
  )

# Data frame for true curve and 1 SD envelope
true_grid <- tibble(x = seq(-5, 5, length.out = 300)) |>
  mutate(
    mu_true = (x^3) / 100,
    lower = mu_true - 0.3,
    upper = mu_true + 0.3
  )

# Base plot on sim_data, explicitly pass true_grid to the annotation layers
gf_point(y ~ x, data = sim_data, alpha = 0.4, size = 1.5) |>
  gf_ribbon(lower + upper ~ x, data = true_grid, fill = "peachpuff", alpha = 0.6, inherit = FALSE) |>
  gf_line(mu_true ~ x, data = true_grid, color = "darkorange", linetype = "dashed", linewidth = 1, inherit = FALSE) |>
  gf_lims(y = c(-3, 3), x = c(-5, 5)) |>
  gf_labs(
    title = "True Data Distribution & Sampled Observations",
    x = "X",
    y = "Y"
  ) |>
  gf_theme(theme_minimal())
```



---

## The Statistical Learning Framework

Let the true data-generating density be $p^*(x, y) = p^*(y \mid x) p^*(x)$ and our parametric candidate model be $q_\theta(x, y) = q_\theta(y \mid x) p^*(x)$.

The Kullback-Leibler (KL) divergence from our model $q_\theta$ to the true distribution $p^*$ is:

$$D_{\text{KL}}(p^* \parallel q_\theta) = \int \int p^*(x, y) \log \frac{p^*(x, y)}{q_\theta(x, y)} \, dy \, dx$$

Expanding the logarithm:

$$D_{\text{KL}}(p^* \parallel q_\theta) = \mathbb{E}_{p^*}\left[\log p^*(X, Y)\right] - \mathbb{E}_{p^*}\left[\log q_\theta(X, Y)\right]$$

Because $\mathbb{E}_{p^*}[\log p^*(X, Y)]$ depends strictly on the true world and is constant with respect to model parameters $\theta$:

$$\arg\min_\theta D_{\text{KL}}(p^* \parallel q_\theta) = \arg\max_\theta \mathbb{E}_{p^*}\left[\log q_\theta(Y \mid X)\right]$$

Under the empirical distribution of a sample $(x_i, y_i)_{i=1}^n \sim p^*$, the empirical expectation replaces the population expectation:

$$\mathbb{E}_{p^*}\left[\log q_\theta(Y \mid X)\right] \approx \frac{1}{n}\sum_{i=1}^n \log q_\theta(y_i \mid x_i)$$

Thus, **Maximum Likelihood Estimation (MLE)** directly approximates **minimal KL divergence** from the truth.

---

## Connecting KL Minimization to Mean Squared Error (MSE)

Consider a conditional Gaussian regression model with fixed variance $\sigma^2$:

$$q_\theta(Y \mid X = x) = \frac{1}{\sqrt{2\pi\sigma^2}} \exp\left( -\frac{(y - f_\theta(x))^2}{2\sigma^2} \right)$$

Evaluating the log-likelihood under this model:

$$\log q_\theta(y_i \mid x_i) = -\frac{1}{2}\log(2\pi\sigma^2) - \frac{1}{2\sigma^2}\left(y_i - f_\theta(x_i)\right)^2$$

Maximizing this objective over $\theta$ discards additive and scaling constants:

$$\arg\min_\theta D_{\text{KL}}(p^* \parallel q_\theta) \iff \arg\min_\theta \frac{1}{n} \sum_{i=1}^n \left(y_i - f_\theta(x_i)\right)^2$$

Minimizing squared error loss is mathematically identical to minimizing the KL divergence under homoscedastic Gaussian assumptions.

---

## Model Fitting: Polynomial Hypothesis Classes

We fit three models via ordinary least squares (equivalent to KL minimization within each model family):

* **Model 1 (Linear):** $\hat{y} = \beta_0 + \beta_1 x$
* **Model 2 (Cubic - Well Specified):** $\hat{y} = \beta_0 + \beta_1 x + \beta_2 x^2 + \beta_3 x^3$
* **Model 3 (Degree 9 - Overparameterized):** $\hat{y} = \sum_{k=0}^9 \beta_k x^k$

```{r}
#| label: model-fitting
mod_linear <- lm(y ~ x, data = sim_data)
mod_cubic  <- lm(y ~ poly(x, 3, raw = TRUE), data = sim_data)
mod_deg9   <- lm(y ~ poly(x, 9, raw = TRUE), data = sim_data)

preds <- true_grid |>
  mutate(
    `Linear (Underfit)` = predict(mod_linear, newdata = true_grid),
    `Cubic (True Form)` = predict(mod_cubic, newdata = true_grid),
    `Degree 9 (High Variance)` = predict(mod_deg9, newdata = true_grid)
  ) |>
  pivot_longer(
    cols = c(`Linear (Underfit)`, `Cubic (True Form)`, `Degree 9 (High Variance)`),
    names_to = "Model",
    values_to = "y_hat"
  )

# Option A: Start with gf_point, then add ribbon and lines explicitly
gf_point(y ~ x, data = sim_data, alpha = 0.25, size = 1.2) |>
  gf_ribbon(lower + upper ~ x, data = true_grid, fill = "peachpuff", alpha = 0.5, inherit = FALSE) |>
  gf_line(y_hat ~ x, data = preds, color = ~Model, linewidth = 1, inherit = FALSE) |>
  gf_lims(y = c(-3, 3), x = c(-5, 5)) |>
  gf_labs(
    title = "Fitted Models via Empirical Risk / KL Minimization",
    subtitle = "Peach band shows true conditional distribution mean ± 1 SD",
    x = "X",
    y = "Y"
  ) |>
  gf_theme(theme_minimal())
```

---

## Empirical KL Divergence Computation

For a Gaussian predictive distribution $q_\theta(y \mid x) = \mathcal{N}(\mu(x), \sigma_\epsilon^2)$ where the true model is $p^*(y \mid x) = \mathcal{N}(\mu^*(x), \sigma^{*2})$, the conditional KL divergence simplifies analytically to:

$$D_{\text{KL}}(p^*(\cdot \mid x) \parallel q_\theta(\cdot \mid x)) = \log\frac{\sigma_\epsilon}{\sigma^*} + \frac{\sigma^{*2} + (\mu^*(x) - \mu(x))^2}{2\sigma_\epsilon^2} - \frac{1}{2}$$

Assuming the estimated noise variance matches $\hat{\sigma}^2 \approx \sigma^{*2} = 0.09$:

$$D_{\text{KL}}(p^*(\cdot \mid x) \parallel q_\theta(\cdot \mid x)) = \frac{(\mu^*(x) - \mu(x))^2}{2\sigma^{*2}}$$

We approximate the expected divergence $\mathbb{E}_X [D_{\text{KL}}]$ across a test set:

```{r}
#| label: kl-evaluation
test_data <- tibble(
  x = runif(10000, -5, 5),
  mu_true = (x^3) / 100
) |>
  mutate(
    pred_linear = predict(mod_linear, newdata = pick(everything())),
    pred_cubic  = predict(mod_cubic, newdata = pick(everything())),
    pred_deg9   = predict(mod_deg9, newdata = pick(everything()))
  )

sigma_sq_true <- 0.09

test_data |>
  summarise(
    KL_Linear = mean((mu_true - pred_linear)^2) / (2 * sigma_sq_true),
    KL_Cubic  = mean((mu_true - pred_cubic)^2)  / (2 * sigma_sq_true),
    KL_Deg9   = mean((mu_true - pred_deg9)^2)   / (2 * sigma_sq_true)
  ) |>
  pivot_longer(everything(), names_to = "Model", values_to = "Estimated_KL") |>
  mutate(Model = str_remove(Model, "KL_")) |>
  knitr::kable(digits = 4, caption = "Expected KL Divergence from True DGP (Lower is Better)")
```

The well-specified cubic model achieves near-zero divergence, whereas model mis-specification (linear) introduces irreducible approximation error (bias), and excessive capacity (degree 9) introduces estimation error (variance).

## Learning Heteroscedastic Representations with a Neural Network

Instead of assuming a fixed variance $\sigma^2$, we can parameterize both the conditional mean and conditional variance using a neural network:

$$z = f(x, \theta) = (z_0, z_1)^\top \in \mathbb{R}^2$$
$$Y \mid Z \sim \text{Normal}\left(\mu = z_0, \, \sigma^2 = g(z_1)^2\right)$$
$$g(z) = 0.01 + \log(1 + e^z) = 0.01 + \text{softplus}(z)$$

### Network Specification
* **Architecture:** $1 \to 32 \to 32 \to 2$
* **Activations:** Continuously Differentiable Exponential Linear Units (`CELU`) for hidden layers
* **Output layer:** 2-dimensional linear representation producing latent variables $z_0$ and $z_1$

### Negative Log-Likelihood as KL Minimization
Under this model, minimizing empirical KL divergence equates to minimizing the negative log-likelihood (NLL):

$$\mathcal{L}(\theta) = \frac{1}{n}\sum_{i=1}^n \left[ \log g(z_{1, i}) + \frac{(y_i - z_{0, i})^2}{2\, g(z_{1, i})^2} \right]$$

```{r}
#| label: neural-network-model
library(torch)

# Set torch seed for reproducibility
torch_manual_seed(42)

# Define Network Architecture
GaussianMLP <- nn_module(
  "GaussianMLP",
  initialize = function() {
    self$net <- nn_sequential(
      nn_linear(1, 32),
      nn_celu(),
      nn_linear(32, 32),
      nn_celu(),
      nn_linear(32, 2)
    )
  },
  forward = function(x) {
    self$net(x)
  }
)

model <- GaussianMLP()
optimizer <- optim_adam(model$parameters, lr = 0.01)

# Prepare tensors
x_train <- torch_tensor(matrix(sim_data$x, ncol = 1), dtype = torch_float())
y_train <- torch_tensor(matrix(sim_data$y, ncol = 1), dtype = torch_float())

# Heteroscedastic Gaussian NLL loss function
gaussian_nll_loss <- function(z, y) {
  mu <- z[, 1, drop = FALSE]
  # g(z1) = 0.01 + softplus(z1)
  sigma <- 0.01 + nnf_softplus(z[, 2, drop = FALSE])
  
  # Negative log likelihood (omitting constant 0.5 * log(2 * pi))
  loss <- torch_log(sigma) + 0.5 * torch_pow((y - mu) / sigma, 2)
  torch_mean(loss)
}

# Training loop
epochs <- 1500
for (epoch in 1:epochs) {
  optimizer$zero_grad()
  z_pred <- model(x_train)
  loss <- gaussian_nll_loss(z_pred, y_train)
  loss$backward()
  optimizer$step()
}
```

### Model Evaluation and Representation Visualization

We compute predictions across the range $x \in [-6, 6]$ along with the conditional standard deviation band $\mu(x) \pm \sigma(x)$.

```{r}
#| label: nn-visualization
# Predict over evaluation grid
eval_grid <- tibble(x = seq(-6, 6, length.out = 400))
x_test_tensor <- torch_tensor(matrix(eval_grid$x, ncol = 1), dtype = torch_float())

model$eval()
with_no_grad({
  z_test <- model(x_test_tensor)
  mu_pred <- as.numeric(z_test[, 1])
  sigma_pred <- as.numeric(0.01 + nnf_softplus(z_test[, 2]))
})

nn_eval <- eval_grid |>
  mutate(
    mu_hat = mu_pred,
    sigma_hat = sigma_pred,
    lower_band = mu_hat - sigma_hat,
    upper_band = mu_hat + sigma_hat
  )

# Calculate conditional KL divergence in bits (log2 base)
# KL(p* || q) = log(sigma_q / sigma_p) + (sigma_p^2 + (mu_p - mu_q)^2)/(2 * sigma_q^2) - 0.5
sigma_true <- sqrt(0.09)
eval_kl <- eval_grid |>
  filter(x >= -5 & x <= 5) |> # Evaluate on support of p*(X)
  mutate(
    mu_p = (x^3) / 100,
    mu_q = approx(nn_eval$x, nn_eval$mu_hat, xout = x)$y,
    sigma_q = approx(nn_eval$x, nn_eval$sigma_hat, xout = x)$y,
    kl_nats = log(sigma_q / sigma_true) + 
      (sigma_true^2 + (mu_p - mu_q)^2) / (2 * (sigma_q^2)) - 0.5,
    kl_bits = kl_nats / log(2)
  )

mean_kl_bits <- round(mean(eval_kl$kl_bits), 2)

# Visualization matching the model diagnostic plot
gf_ribbon(lower + upper ~ x, data = true_grid, fill = "peachpuff", alpha = 0.5) |>
  gf_line(mu_true ~ x, data = true_grid, color = "darkorange", linetype = "dashed", linewidth = 0.8) |>
  gf_ribbon(lower_band + upper_band ~ x, data = nn_eval, fill = "#1f77b4", alpha = 0.35, inherit = FALSE) |>
  gf_line(mu_hat ~ x, data = nn_eval, color = "#1f77b4", linewidth = 1.1, inherit = FALSE) |>
  gf_lims(y = c(-3, 3), x = c(-6, 6)) |>
  gf_labs(
    title = "Neural Network Representation & Uncertainty Band",
    subtitle = paste0("Mean conditional KL divergence: ", mean_kl_bits, " bits"),
    x = "X",
    y = "Y"
  ) |>
  gf_theme(theme_minimal())
```

